% =============================================================
% Theory Companion Paper — LaTeX skeleton (T0-2, WOV-261)
% =============================================================
% Working title (charter §0):
%   Structured Representation for Code LLMs: When Format Can and Cannot
%   Help a Bounded Learner, and Why.
%
% This is a SECTION-STUB skeleton. Each section is a separate file under
% sections/ carrying a 1-3 sentence stub plus a pointer to the charter
% milestone that fills it. Full mathematics is deferred to T1/T2.
%
% Authority: subordinate to docs/design-constitution.tex and the frozen
% preprint-v1. Governed by docs/theory-paper-charter.md. The honesty
% guardrails (charter §3, G1-G6) are binding on every section.
%
% Build: see paper-theory/README.md or `make theory-paper`.
% Primary tool: tectonic (self-contained, resolves natbib automatically).

\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}        % \vcentcolon, \coloneqq helpers used in notation
\usepackage{bm}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage[round]{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage[capitalize, noabbrev]{cleveref}

% ----- Theorem environments (numbered within section) -----
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
% ----- Falsifiable predictions (§8, T3-2) -----
% Predictions are NOT proved results: they carry an independent counter (P8.x)
% so they are never confused with the theorems they derive from, and they live
% in remark style to keep the conjectural status visually distinct (guardrails
% G2/G6). Each prediction is stated with a sign/size claim and an explicit
% falsifier in its body.
\newtheorem{prediction}{Prediction}[section]

% ----- Idealization-disclosure box (guardrail G3) -----
% Every invocation of the irrelevance theorem must state which idealization
% it relies on. This command standardizes that disclosure.
\newcommand{\idealization}[1]{%
  \par\vspace{2pt}\noindent\textit{Idealization disclosure (G3).} #1\par\vspace{2pt}}

% ----- Go/no-go marker (charter §2 discipline) -----
% Sections 3 and 6 are provisional: they survive only if they earn a
% theorem. This banner keeps that status visible in the draft.
\newcommand{\gonogo}[1]{%
  \par\noindent\fbox{\parbox{0.97\linewidth}{\small\textbf{GO/NO-GO (charter §2).}
  #1}}\par\vspace{4pt}}

\input{notation}

\title{Structured Representation for Code LLMs:\\
       When Format Can and Cannot Help a Bounded Learner, and Why}

\author{%
  \textbf{\large Takumi Shiraishi} \\[2pt]
  Independent Researcher \\[1pt]
  \href{mailto:takumi@wovol.com}{\texttt{takumi@wovol.com}}
}

\date{Draft skeleton --- \today}

\begin{document}
\maketitle

\begin{abstract}
% Drafted in T4-1 after the theorems exist; backbone is charter §1.
Does the format in which code is presented to a language model change what the
model can predict about it? We give a theoretical companion to the frozen
empirical record of \texttt{preprint-v1}, whose central pre-registered cell
returned a clean null. Making representations measurable maps and predictors
Markov kernels on a shared standard Borel space, we prove a two-halved answer.
\emph{Negatively}, under \emph{information parity} --- two representations
retaining the same mutual information with the target --- the data-processing
inequality closes at the Bayes optimum and the optimal risk is
format-invariant, so the recorded null is the \emph{predicted} information-limit
outcome rather than an anomaly; when parity fails, the inflation of Bayes risk
is controlled by the information lost (exactly under log loss, within a
square-root envelope otherwise). \emph{Positively} --- and never as a claim that
any effect exists --- a finite-capacity learner does not attain the information
limit, and the residual is exactly where format may act: at constant
information the achievable representation advantage is capped by a capacity
budget carrying no mutual-information term, two-sided in magnitude, and
non-vacuous only in a named finite-resource, capacity-separated regime. Two
go/no-go refinements show operational parity is a retraction whose round-trip
defect is invisible to both the mutual information and the Bayes risk yet
dominates the swap gap of any fixed bounded predictor. We map the three frozen
Lumen outcomes to the regimes they occupy --- claiming consistency, not
confirmation --- and convert the localization into five falsifiable predictions
on the same apparatus, each with an explicit falsifier. The thesis: format
cannot supply information that is not there, so any genuine
code-representation effect on a bounded model is a capacity phenomenon of
bounded size, appearing only near the model's frontier.
\end{abstract}

\tableofcontents
\clearpage

\input{sections/01-introduction}
\input{sections/02-preliminaries}
\input{sections/03-category-of-representations}
\input{sections/04-information-parity-irrelevance}
\input{sections/05-bounded-learner-gap}
\input{sections/06-parity-defect}
\input{sections/07-empirical-anchoring}
\input{sections/08-predictions}
\input{sections/09-related-work}
\input{sections/10-limitations-honesty}
\input{sections/11-conclusion}

\appendix
\input{sections/appendix-proofs}

\bibliographystyle{plainnat}
\bibliography{theory}

\end{document}
