% Section 4 — Information parity and the irrelevance theorem
% (charter §4.4; issues T1-2, T2-1, T2-2)
\section{Information parity and the irrelevance theorem}
\label{sec:parity}

% Area role (charter §2): parity = sufficiency / Markov-equivalence; the
% irrelevance theorem via DPI; rate-distortion for lossy reps.
%
% This file holds:
%   - T1-2 (WOV-263): information-theoretic foundations (entropy, divergence,
%     mutual information), the data-processing inequality with proof,
%     sufficiency + Fisher-Neyman factorization, rate-distortion for lossy
%     reps, and the FORMAL DEFINITION of information parity with the
%     operational-vs-absolute distinction (paper §3.3) made precise.
%   - T2-1 (WOV-267): structured representation + decoding relation
%     (subsec:structured), the C-ladder worked examples (tab:cladder), and the
%     parity-hypothesis ledger naming which grade each later theorem assumes
%     (tab:parity-ledger).
%   - T2-2: the at-optimum irrelevance theorem (DPI => Bayes-risk invariance).

This section supplies the information-theoretic engine of the paper. It builds
the three quantities the spine needs (entropy, relative entropy, mutual
information), proves the data-processing inequality in the exact form the
irrelevance theorem consumes, identifies \emph{sufficiency} as the structural
condition under which the inequality is tight, and then defines
\emph{information parity} as that tightness condition. The negative half of
the thesis (\cref{sec:intro}) is, at this level, a single sentence: parity
makes the data-processing inequality an equality, so the information available
to an optimal predictor does not move with the representation. Making that
sentence a theorem about Bayes risk is deferred to T2-2; here we fix the
objects it quantifies over. We follow the standard treatments of
\citet{coverthomas2006} for the information quantities and rate--distortion,
and \citet{lehmanncasella1998} for sufficiency and the factorization theorem;
all objects are the measure-theoretic ones of \cref{sec:prelim}.

\subsection{Entropy, relative entropy, mutual information}
\label{subsec:infoquantities}

Throughout, $X \markov \rep(X) \markov Y$ and related Markov statements are
read against the joint law $\jlaw$ of \cref{ass:sbs}; all random elements live
on the common space $(\Omega,\mathcal{F},\Prob)$. We state the discrete forms
(finite token alphabets and countable program syntax are the operative case
for \texttt{preprint-v1}); the definitions extend to the standard Borel
setting of \cref{sec:prelim} by taking suprema over finite measurable
partitions \citep[Ch.~8]{coverthomas2006}, and nothing downstream depends on
the continuous refinement.

\begin{definition}[Entropy and conditional entropy]
\label{def:entropy}
For a discrete random element $U$ with law $p_U$, the \emph{entropy} is
$\Ent{U} \defeq -\sum_{u} p_U(u)\log p_U(u)$, and for a pair $(U,V)$ the
\emph{conditional entropy} is
$\CondEnt{U}{V} \defeq -\sum_{u,v} p_{UV}(u,v)\log p_{U \given V}(u\given v)$,
with the convention $0\log 0 = 0$. Logarithms are natural; entropies are in
nats.
\end{definition}

\begin{definition}[Relative entropy]
\label{def:kl}
For probability measures $\mu \ll \nu$ on a common space, the
\emph{relative entropy} (Kullback--Leibler divergence) is
$\KL{\mu}{\nu} \defeq \int \log\!\big(\rnd{\mu}{\nu}\big)\,\mathrm{d}\mu$,
with $\KL{\mu}{\nu} = +\infty$ when $\mu \not\ll \nu$. Gibbs' inequality gives
$\KL{\mu}{\nu} \ge 0$, with equality iff $\mu = \nu$.
\end{definition}

\begin{definition}[Mutual information]
\label{def:mi}
The \emph{mutual information} between $U$ and $V$ is the relative entropy
between their joint law and the product of the marginals,
\[
  \MI{U}{V} \defeq \KL{\law{(U,V)}}{\,\law{U}\otimes\law{V}\,}
            = \Ent{U} - \CondEnt{U}{V}
            = \Ent{V} - \CondEnt{V}{U}.
\]
By \cref{def:kl}, $\MI{U}{V} \ge 0$, with equality iff $U \indep V$. The
\emph{conditional mutual information} $\MI{U}{V \given W}$ is defined
analogously from the conditional laws, and $\MI{U}{V\given W}=0$ iff
$U \indep V \given W$.
\end{definition}

The one identity we use repeatedly is the \emph{chain rule}: for any $U,V,W$,
\begin{equation}
  \label{eq:chainrule}
  \MI{U}{(V,W)} \;=\; \MI{U}{V} \;+\; \MI{U}{W \given V},
\end{equation}
which follows by writing each term via \cref{def:entropy} and cancelling
\citep[Thm.~2.5.2]{coverthomas2006}.

\subsection{The data-processing inequality}
\label{subsec:dpi}

The inequality is stated for a Markov chain and then specialized to a
representation, which is the only instance the paper uses.

\begin{lemma}[Data-processing inequality]
\label{lem:dpi}
If $U \markov V \markov W$ is a Markov chain (i.e.\ $U \indep W \given V$),
then $\MI{U}{W} \le \MI{U}{V}$, with equality iff $U \markov W \markov V$ also
holds (i.e.\ $V$ carries no information about $U$ beyond what $W$ already does).
\end{lemma}

\begin{proof}
Expand $\MI{U}{(V,W)}$ by the chain rule \eqref{eq:chainrule} in the two
orders:
\[
  \MI{U}{V} + \MI{U}{W \given V}
  \;=\; \MI{U}{(V,W)} \;=\;
  \MI{U}{W} + \MI{U}{V \given W}.
\]
The Markov hypothesis $U \indep W \given V$ is exactly $\MI{U}{W\given V}=0$
(\cref{def:mi}), so $\MI{U}{V} = \MI{U}{W} + \MI{U}{V \given W}$. Since
$\MI{U}{V\given W}\ge 0$, we obtain $\MI{U}{V} \ge \MI{U}{W}$. Equality holds
iff $\MI{U}{V\given W}=0$, i.e.\ $U \indep V \given W$, which is the Markov
chain $U \markov W \markov V$.
\end{proof}

\begin{corollary}[A representation cannot create predictive information]
\label{cor:dpi-rep}
For any representation $\rep$ of \cref{def:rep}, $\rep(X)$ is a measurable
function of $X$, so $Y \markov X \markov \rep(X)$ is a Markov chain. Hence
\begin{equation}
  \label{eq:dpi-rep}
  \MI{Y}{\rep(X)} \;\le\; \MI{Y}{X},
\end{equation}
with equality iff $Y \markov \rep(X) \markov X$, i.e.\ iff $\rep(X)$ retains
all of $X$'s information about $Y$.
\end{corollary}

\begin{proof}
$\rep(X)$ is $\sigalg{X}$-measurable, so it is conditionally independent of
$Y$ given $X$: $Y \indep \rep(X) \given X$, which is the chain
$Y \markov X \markov \rep(X)$. Apply \cref{lem:dpi} with
$(U,V,W)=(Y,X,\rep(X))$.
\end{proof}

\cref{cor:dpi-rep} is the formal content of \cref{rem:lossy} and of the
monotonicity noted after \cref{def:bayesrisk}: a representation can only
coarsen the input $\sigma$-algebra, hence can only weakly decrease the
information about $Y$. The equality case is the object the rest of the section
names.

\subsection{Sufficiency and the Fisher--Neyman factorization}
\label{subsec:sufficiency}

The equality case of \cref{cor:dpi-rep} is classical sufficiency, transported
from parameter estimation to prediction by reading $Y$ in the role of the
parameter.

\begin{definition}[Sufficiency for $Y$]
\label{def:sufficiency}
A representation $\rep$ is \emph{sufficient for $Y$} if
$Y \indep X \given \rep(X)$, equivalently if the Markov chain
$Y \markov \rep(X) \markov X$ holds. We write $\rep \suff \mathrm{id}$ for
``$\rep$ is sufficient'' and, more generally, $\rep \suff \rep'$ when
$\sigalg{\rep'} \subseteq \sigalg{\rep}$ up to $\Prob$-null sets and $\rep$ is
sufficient relative to $\rep'$.
\end{definition}

\begin{proposition}[Sufficiency $\Leftrightarrow$ information equality]
\label{prop:suff-iff-mi}
$\rep$ is sufficient for $Y$ if and only if $\MI{Y}{\rep(X)} = \MI{Y}{X}$.
\end{proposition}

\begin{proof}
By \cref{cor:dpi-rep}, equality in \eqref{eq:dpi-rep} holds iff
$Y \markov \rep(X) \markov X$, which is \cref{def:sufficiency}.
\end{proof}

\begin{proposition}[Fisher--Neyman factorization]
\label{prop:fisher-neyman}
Suppose the conditional law of $X$ given $Y=y$ has density $f_y$ with respect
to a common $\sigma$-finite measure $\lambda$ on $\Xspace$. Then $\rep$ is
sufficient for $Y$ if and only if there exist measurable
$g_y : \Zspace_{\rep} \to [0,\infty)$ and $h : \Xspace \to [0,\infty)$ with
\begin{equation}
  \label{eq:fisher-neyman}
  f_y(x) \;=\; g_y\!\big(\rep(x)\big)\, h(x)
  \qquad \lambda\text{-a.e., for every } y.
\end{equation}
\end{proposition}

\begin{proof}
This is the factorization theorem \citep[Thm.~1.6.5]{lehmanncasella1998} with
the parameter index identified with the target value $y$ and the statistic
identified with the representation map $\rep$. The standard-Borel hypothesis
of \cref{ass:sbs} supplies the regular conditional laws the theorem requires.
\end{proof}

\cref{prop:fisher-neyman} is the operational test: a representation is
information-lossless for $Y$ exactly when the data likelihood depends on the
input only through $\rep$. The dependence on $x$ that factors into $h(x)$ is
the part a representation may freely discard --- it is the formal image of
``format that carries no predictive content.''

\subsection{Rate--distortion for lossy representations}
\label{subsec:ratedist}

When a representation is \emph{not} sufficient, \cref{cor:dpi-rep} is strict
and some predictive information is lost. The rate--distortion function
\citep[Ch.~10]{coverthomas2006} measures the minimal information a
representation must retain to keep a fidelity constraint, and converts the
information gap into a bound on achievable risk.

\begin{definition}[Rate--distortion function]
\label{def:ratedist}
For a fidelity (distortion) measure $d : \Xspace\times\Zspace_{\rep} \to
[0,\infty)$, the \emph{rate--distortion function} at level $D$ is
\[
  \ratedist \;\defeq\;
  \inf_{\,\rep \,:\, \Expect[d(X,\rep(X))]\le D}\, \MI{X}{\rep(X)},
\]
the infimum over (possibly stochastic) representations meeting the
distortion budget $D$.
\end{definition}

\begin{remark}[Lossy parity defect, previewed]
\label{rem:lossy-defect}
A representation that is sufficient for $Y$ sits at $D=0$ on the relevant
$Y$-fidelity axis: it loses no predictive information and \eqref{eq:dpi-rep}
is an equality. A lossy representation incurs a strictly positive
information gap $\MI{Y}{X}-\MI{Y}{\rep(X)} > 0$, and the rate--distortion
machinery bounds the resulting inflation of Bayes risk. The exact bound is the
rate--distortion corollary of the irrelevance theorem (T2-2); its quantitative
companion --- a Wasserstein/Fisher \emph{parity defect} computed against the
R4 round-trip data --- is \cref{sec:parity-defect}. We record here only that
the defect is governed by an information gap, so that ``how far from parity''
is a measurable quantity rather than a metaphor.
\end{remark}

\subsection{Information parity: operational versus absolute}
\label{subsec:parity-def}

We can now define the central object. Two grades are distinguished, because
the distinction is exactly guardrail~G3's ``parity is operational, not
absolute'' (paper §3.3), and the paper uses the weaker grade deliberately.

\begin{definition}[Information parity]
\label{def:parity}
Two representations $\rep, \rep'$ are in \emph{(operational) information
parity for $Y$}, written $\rep \parityeq \rep'$, if both are sufficient for
$Y$ in the sense of \cref{def:sufficiency}; equivalently, by
\cref{prop:suff-iff-mi},
\begin{equation}
  \label{eq:parity}
  \MI{Y}{\rep(X)} \;=\; \MI{Y}{\rep'(X)} \;=\; \MI{Y}{X}.
\end{equation}
They are in \emph{absolute parity} if moreover $\sigalg{\rep} = \sigalg{\rep'}$
up to $\Prob$-null sets (their induced sub-$\sigma$-algebras agree, so each is
a measurable function of the other almost surely).
\end{definition}

\begin{proposition}[Absolute parity is strictly stronger]
\label{prop:abs-strict}
Absolute parity implies operational parity. The converse fails: there exist
$\rep \parityeq \rep'$ with $\sigalg{\rep} \ne \sigalg{\rep'}$.
\end{proposition}

\begin{proof}
If $\sigalg{\rep}=\sigalg{\rep'}$ then each is a measurable function of the
other, so $\MI{Y}{\rep(X)}=\MI{Y}{\rep'(X)}$ and both equal $\MI{Y}{X}$ when
either is sufficient; this gives \eqref{eq:parity}. For the converse, let
$\rep' = (\rep, N)$ adjoin a component $N \indep (X,Y)$ that is independent
noise: $\rep'$ retains every predictive bit $\rep$ does, so
$\MI{Y}{\rep'(X)}=\MI{Y}{\rep(X)}$ and $\rep\parityeq\rep'$, yet
$\sigalg{\rep'} \supsetneq \sigalg{\rep}$. More to the point for the C-ladder:
$\ladderplus$ and $\ladder{4}$ can carry the same information about $Y$ while
generating different $\sigma$-algebras over byte-level inputs, which is
precisely what the round-trip data witness.
\end{proof}

\begin{remark}[Which grade the paper uses, and why it matters]
\label{rem:which-parity}
The irrelevance theorem (T2-2) is stated against \emph{operational} parity
(\cref{def:parity}): that is all the Bayes-risk argument needs, since
\cref{prop:suff-iff-mi} ties operational parity to the equality case of the
data-processing inequality, and Bayes risk depends on a representation only
through the predictive information it retains (\cref{prop:disint}). Absolute
parity is sufficient but unnecessarily strong, and --- by
\cref{prop:abs-strict} --- it is the grade that \emph{fails} in practice: the
control $\ladderplus$ and the structured representation $\ladder{4}$ are
designed to be operationally parity (equal predictive content for $Y$) but are
not byte-identical, and the R4 round-trip recovery (\cref{sec:parity-defect})
quantifies the residual asymmetry $\sigalg{\ladder{4}} \ne
\sigalg{\ladderplus}$. The slogan ``parity is operational, not absolute'' is
therefore the formal statement that the central comparison rests on
\eqref{eq:parity} and not on $\sigma$-algebra equality; the gap between the
two grades is real, bounded, and quantified in \cref{sec:parity-defect}, never
assumed away.
\idealization{Operational parity \eqref{eq:parity} is an equality of mutual
informations under the (unknown but fixed) data law $\jlaw$. It is an
idealization in two senses made explicit downstream: (i) it presumes the two
representations are \emph{exactly} sufficient, whereas a constructed control
may be sufficient only up to a measured defect (\cref{sec:parity-defect}); and
(ii) the risk consequence it feeds (T2-2) presumes Bayes-optimality, which a
finite-capacity LLM does not attain (\cref{sec:bounded}).}
\end{remark}

\subsection{Structured representations and the C-ladder}
\label{subsec:structured}

\cref{def:parity} compares arbitrary measurable maps. The paper's actual
objects are narrower: a code representation does not merely coarsen the input,
it \emph{re-renders} it, adding derived structure (a parse tree, inferred
types, extracted contracts) while remaining, by design, decodable back toward
source form. That decodability is what makes $\ladderplus$ a usable control
rather than an abstract coarsening, and it is the hinge on which operational
and absolute parity (\cref{def:parity}) separate. We make the object precise
here (T2-1), then instantiate it on the C-ladder, and close by naming, for
each later theorem, the exact parity grade it assumes.

\begin{definition}[Structured representation and its decoding relation]
\label{def:structured-rep}
A \emph{structured representation} is a representation $\rep$ (\cref{def:rep})
equipped with a \emph{decoding relation}: a Markov kernel
$\decode_{\rep} : \Zspace_{\rep}\times\Bor{\Xspace}\to[0,1]$ (\cref{def:kernel})
that attempts to recover the input from its encoding. We write the pair as
$\strrep$. Two faithfulness grades are distinguished.
\begin{enumerate}
  \item[(i)] $\strrep$ is \emph{exactly faithful} (input-lossless) if the
    decoding recovers $X$ almost surely, i.e.\ $\decode_{\rep}(\rep(X),\cdot)$
    is the point mass at $X$ ($\Prob$-a.s.); equivalently $\rep$ is injective
    up to $\Prob_X$-null sets, so $\repsig = \Bor{\Xspace}$ and $\rep$ is
    invertible on a set of full measure.
  \item[(ii)] $\strrep$ is \emph{$Y$-faithful} (lossless for $Y$) if the
    decoding recovers the predictive content of $X$ for $Y$: the posterior is
    unchanged under encode--decode,
    $\condexp{\mathbf{1}_{\{Y\in B\}}}{\rep(X)} =
     \condexp{\mathbf{1}_{\{Y\in B\}}}{X}$ $\Prob$-a.s.\ for every
    $B\in\Bor{\Yspace}$. By \cref{prop:disint} this is exactly sufficiency of
    $\rep$ for $Y$ (\cref{def:sufficiency}), hence by \cref{prop:suff-iff-mi}
    exactly $\MI{Y}{\rep(X)}=\MI{Y}{X}$.
\end{enumerate}
A structured representation is \emph{lossy} (for $Y$) when (ii) fails.
\end{definition}

The decoding relation is the measure-theoretic shadow of the re-parse/project
morphisms of \cref{sec:category}: $\decode_{\rep}$ is the kernel form of the
ladder morphism $p:\ladder{4}\to\ladderplus$ that \cref{def:cat-parity} reads
as a retraction. Reading \cref{def:structured-rep} against
\cref{def:retract,def:cat-parity} lines the two languages up exactly: exact
faithfulness is the \emph{isomorphism} case ($\decode_{\rep}$ inverts $\rep$ on
both sides), while $Y$-faithfulness is the weaker \emph{retraction} case
($\decode_{\rep}$ recovers only the part of $X$ that the posterior depends on).
This is the same operational-versus-absolute split, now carried by an explicit
decoder rather than by a $\sigma$-algebra comparison.

\begin{proposition}[Decoding faithfulness fixes the parity grade]
\label{prop:decode-parity}
Let $\strrep$ and $\langle\rep',\decode_{\rep'}\rangle$ be structured
representations.
\begin{enumerate}
  \item[(a)] If both are $Y$-faithful, then $\rep\parityeq\rep'$
    (operational parity, \cref{def:parity}).
  \item[(b)] If both are exactly faithful, then $\sigalg{\rep}=\sigalg{\rep'}
    =\Bor{\Xspace}$ up to $\Prob$-null sets, so $\rep,\rep'$ are in absolute
    parity.
  \item[(c)] The converse of (a) fails: there exist $Y$-faithful $\rep\parityeq
    \rep'$ that are not exactly faithful, with $\sigalg{\rep}\ne\sigalg{\rep'}$.
\end{enumerate}
\end{proposition}

\begin{proof}
(a)~$Y$-faithfulness of each is, by \cref{def:structured-rep}(ii) and
\cref{prop:suff-iff-mi}, the equality $\MI{Y}{\rep(X)}=\MI{Y}{X}$ and likewise
for $\rep'$; this is \eqref{eq:parity}, i.e.\ $\rep\parityeq\rep'$. (b)~Exact
faithfulness gives $\repsig=\Bor{\Xspace}=\sigalg{\rep'}$ up to null sets by
\cref{def:structured-rep}(i), which is absolute parity (\cref{def:parity}).
(c)~Take $\rep$ exactly faithful and $\rep'=\rep$ post-composed with the
forgetful map that erases a predictively inert coordinate of $X$ (one
$\indep Y$ given the rest); $\rep'$ is $Y$-faithful but not exactly faithful,
and $\sigalg{\rep'}\subsetneq\sigalg{\rep}$. This is the structured-rep
instance of \cref{prop:abs-strict}, and the $\ladderplus/\ladder{4}$ pair below
is its concrete witness.
\end{proof}

\paragraph{Worked examples: the C-ladder.}
The rungs of \texttt{preprint-v1} (Constitution §14.1) are structured
representations in the sense of \cref{def:structured-rep}; each is a
deterministic measurable function of the plain source, so by
\cref{cor:dpi-rep} none can \emph{increase} $\MI{Y}{\cdot}$, and the only
question is whether each is $Y$-faithful (lossless) or merely close to it.
\cref{tab:cladder} records the encode/decode pair and both faithfulness grades.

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}llll@{}}
\toprule
Rung & Structure added & Decoding $\decode$ & Faithfulness \\
\midrule
$\ladder{1}$ (plain source) & --- (baseline)        & identity            & exact \\
$\ladder{2}$ (Raw AST)      & parse tree            & unparse / print     & $Y$-faithful, not exact \\
$\ladder{3}$ (Typed AST)    & inferred types        & type-erase $+$ unparse & $Y$-faithful, not exact \\
$\ladder{4}$ (Contract IR)  & extracted contracts   & project to source   & $Y$-faithful, not exact \\
$\ladderplus$ (control)     & annotations from $\ladder{4}$ & re-parse to IR & $Y$-faithful, not exact \\
\bottomrule
\end{tabular}
\caption{The C-ladder as structured representations (\cref{def:structured-rep}).
Each rung is a measurable function of $\ladder{1}$, so by \cref{cor:dpi-rep} it
cannot create predictive information. Surface bytes (whitespace, comment
layout) live in the part the decoders discard; the design intent is that this
discarded part is predictively inert, i.e.\ each rung is $Y$-faithful. ``Not
exact'' marks the rungs whose decoder does \emph{not} recover the input
byte-for-byte, so they fail exact faithfulness and hence absolute parity.}
\label{tab:cladder}
\end{table}

The control pair is the load-bearing row. $\ladderplus$ is the projection of
$\ladder{4}$ to annotated source (\cref{subsec:catrep}); by construction it
carries the same contracts, types, and node kinds as $\ladder{4}$, so the two
are designed to be \emph{$Y$-faithful} and hence in operational parity
$\ladderplus\parityeq\ladder{4}$ (\cref{prop:decode-parity}(a)). They are
\emph{not} exactly faithful: the round trip
$\roundtrip{\ladder{4}}{\ladderplus}$ recovers contracts at $100\%$, node kinds
at $99.8\%$, and types at $96.4\%$, but byte-exact ASTs at $0/30$
(R4 data, \cref{sec:parity-defect}). So $\sigalg{\ladder{4}}\ne
\sigalg{\ladderplus}$, absolute parity fails, and the residual asymmetry is not
assumed away but measured in \cref{sec:parity-defect}. This is precisely the
configuration \cref{prop:decode-parity}(c) describes.

\begin{remark}[Which parity grade each theorem assumes]
\label{rem:parity-ledger}
The point of separating the grades is that the paper's results do not all
need the same one. \cref{tab:parity-ledger} is the ledger; it is binding, in
the sense that no later theorem may silently upgrade its hypothesis from
operational to absolute parity (that upgrade is exactly the move
guardrail~G3 forbids).

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lll@{}}
\toprule
Result & Parity hypothesis & Role of the hypothesis \\
\midrule
\cref{cor:dpi-rep} (DPI for reps)            & none           & holds for every representation \\
\cref{thm:irrelevance} (T2-2, irrelevance)   & operational    & $\Rightarrow$ equal Bayes risk \\
\S\ref{sec:bounded} bound (T2-4, \cref{def:advantage}) & operational & information held equal; gap is capacity \\
\S\ref{sec:category} retraction (T2-3, \cref{thm:retraction}) & operational $=$ retraction & defect $=$ idempotent, info-invisible \\
\S\ref{sec:parity-defect} defect (T2-5, \cref{def:parity-defect}) & operational holds & quantifies failure of \emph{absolute} \\
\bottomrule
\end{tabular}
\caption{Parity-hypothesis ledger. Every downstream result is stated against
\emph{operational} parity (\cref{def:parity}, \eqref{eq:parity}); absolute
parity is never assumed, only either characterized (\S\ref{sec:category}) or
measured as a defect (\S\ref{sec:parity-defect}).}
\label{tab:parity-ledger}
\end{table}
\idealization{Every row that assumes operational parity inherits the two
idealizations of \cref{rem:which-parity}: exact sufficiency of the compared
representations (relaxed to a measured defect in \cref{sec:parity-defect}) and,
for \cref{thm:irrelevance}, Bayes-optimality (relaxed in \cref{sec:bounded}).
The ledger's value is that it localizes \emph{where} each idealization is
spent, so no single theorem carries more than it states.}
\end{remark}

\subsection{The irrelevance theorem}
\label{subsec:irrelevance}

With parity defined (\cref{def:parity}) and the data-processing inequality in
hand (\cref{lem:dpi,cor:dpi-rep}), the at-optimum invariance is now a theorem:
operational parity \eqref{eq:parity} forces the Bayes risks to coincide. This
is the load-bearing result of the paper --- the formal statement that makes the
\texttt{preprint-v1} T1 null a \emph{prediction} (charter~G1) rather than a
disappointment. The argument is the one previewed after \cref{def:bayesrisk}:
sufficiency makes the regular conditional law of $Y$ given the representation
agree with the law given the raw input (\cref{prop:disint}), and Bayes risk is
a functional of that conditional law; equal posteriors therefore force equal
Bayes risk. The proof is carried out in full in \cref{app:irrelevance}; we
state the result and its rate--distortion corollary here, each with its
idealization disclosure.

\begin{theorem}[Irrelevance at optimum]
\label{thm:irrelevance}
Let $\loss$ be a loss admitting a measurable Bayes act (so that
\cref{prop:disint} applies), and let $\rep,\rep'$ be representations in
operational information parity for $Y$, i.e.\ $\rep \parityeq \rep'$
(\cref{def:parity}). Then the Bayes risk is format-invariant:
\begin{equation}
  \label{eq:irrelevance}
  \bayesrisk(\rep) \;=\; \bayesrisk(\rep') \;=\; \bayesrisk(\mathrm{id}).
\end{equation}
\end{theorem}

\begin{remark}[The precise slogan]
\label{rem:slogan}
\cref{thm:irrelevance} is the exact content of ``at constant information, the
optimum cannot distinguish format.'' Each clause is now a named object:
\emph{constant information} is the parity equality \eqref{eq:parity},
$\MI{Y}{\rep(X)} = \MI{Y}{\rep'(X)} = \MI{Y}{X}$; \emph{the optimum} is the
Bayes predictor $\post{\rep}$ of \cref{prop:disint}, the infinite-capacity
posterior and not any realizable learner; \emph{cannot distinguish format} is
\eqref{eq:irrelevance}, equality of the information-limit risks. The theorem is
silent on which representation a \emph{bounded} predictor prefers: that is the
whole subject of \cref{sec:bounded}, and this negative result is exactly what
makes the positive question there well posed rather than confounded by an
information difference.
\idealization{\eqref{eq:irrelevance} holds at two idealizations, neither met by
a frozen LLM. \emph{(i)~Bayes-optimality.} $\bayesrisk(\rep)$ is the
infinite-capacity information limit (\cref{def:bayesrisk}), attained only by the
posterior $\post{\rep}$; a finite-capacity predictor does not reach it, and the
residual gap is precisely where format may act (\cref{sec:bounded}).
\emph{(ii)~Exact parity.} \eqref{eq:parity} is an exact equality of mutual
informations under the fixed but unknown data law $\jlaw$, whereas a constructed
control attains it only up to a measured defect: $\ladderplus$ and $\ladder{4}$
are designed for operational parity but are not byte-identical
(\cref{prop:abs-strict}), and \cref{sec:parity-defect} quantifies the residual.
By the ledger (\cref{tab:parity-ledger}) every downstream invocation of this
theorem inherits these two idealizations and no others.}
\end{remark}

When a representation is \emph{not} sufficient, \eqref{eq:irrelevance} fails,
but not arbitrarily: the inflation of Bayes risk is controlled by the
information the representation loses. This is the rate--distortion reading
promised in \cref{rem:lossy-defect}, and it is what turns ``how far from
parity'' into a bound on ``how far from optimal.''

\begin{corollary}[Rate--distortion bound on the parity defect]
\label{cor:ratedist-gap}
For a representation $\rep$ write its \emph{information gap} as
\[
  \infogap{\rep} \;\defeq\; \MI{Y}{X} - \MI{Y}{\rep(X)} \;\ge\; 0,
\]
which is $0$ iff $\rep$ is sufficient for $Y$ (\cref{prop:suff-iff-mi}). The
Bayes risk inflates monotonically with the gap, $\bayesrisk(\rep) \ge
\bayesrisk(\mathrm{id})$, and the inflation is bounded as follows.
\begin{enumerate}
  \item[(a)] \emph{(Logarithmic loss --- exact.)} For the log loss
    $\loss(y,q) = -\log q(y)$ over predictive distributions
    $q \in \Aspace = \mathcal{P}(\Yspace)$ on a discrete $\Yspace$, the Bayes
    risk is conditional entropy and the inflation \emph{equals} the lost
    information:
    \begin{equation}
      \label{eq:logloss-exact}
      \bayesrisk(\rep) - \bayesrisk(\mathrm{id})
      \;=\; \CondEnt{Y}{\rep(X)} - \CondEnt{Y}{X}
      \;=\; \infogap{\rep}.
    \end{equation}
  \item[(b)] \emph{(Bounded loss --- general.)} For any loss with
    $\|\loss\|_\infty \defeq \sup_{y,a}\loss(y,a) < \infty$,
    \begin{equation}
      \label{eq:ratedist-bound}
      0 \;\le\; \bayesrisk(\rep) - \bayesrisk(\mathrm{id})
        \;\le\; \|\loss\|_\infty\,\sqrt{\tfrac12\,\infogap{\rep}}.
    \end{equation}
\end{enumerate}
In particular the inflation is continuous in the information gap and vanishes as
$\infogap{\rep}\to 0$; parity is the boundary case $\infogap{\rep}=0$ at which
both bounds collapse to the equality \eqref{eq:irrelevance}.
\end{corollary}

\begin{remark}[Reading the corollary, and its handoff to \cref{sec:parity-defect}]
\label{rem:ratedist-reading}
Identity \eqref{eq:logloss-exact} is the rate--distortion statement in its
sharpest form: under the canonical proper loss the achievable Bayes risk is
exactly the conditional entropy, so the distortion floor and the retained rate
$\MI{Y}{\rep(X)}$ move together, and the parity boundary $D=0$ of
\cref{rem:lossy-defect} is the sufficiency boundary $\infogap{\rep}=0$. Bound
\eqref{eq:ratedist-bound} is the general guarantee --- a representation that
loses little predictive information cannot inflate the optimal risk by much,
whatever the bounded loss. Both are statements about the \emph{optimum}, and
both reach equality at parity. The residual they cannot see is the one
\cref{sec:parity-defect} measures: two representations with
$\infogap{\rep}=\infogap{\rep'}=0$ --- exact information parity, zero inflation
here --- may still drive a \emph{bounded} predictor to different predictive
laws, and the parity defect $\paritydefect$ \eqref{eq:parity-defect} is exactly
that bounded-regime residual, invisible at the optimum by
\cref{thm:irrelevance}.
\idealization{\cref{cor:ratedist-gap} relaxes idealization~(ii) of
\cref{rem:slogan} --- it is the lossy, off-parity case --- but retains~(i): it
bounds the inflation of the \emph{Bayes} risk, not of a bounded predictor's
risk. The information gap $\infogap{\rep}$ is an exact quantity whose zero set
is characterized by the factorization \cref{prop:fisher-neyman}; converting it
into a \emph{bounded}-predictor risk gap is not done here and is the open
obligation handed to \cref{sec:parity-defect} (T2-5) under its go/no-go.}
\end{remark}
