% Section 5 — The bounded-learner gap (charter §4.5; issues T1-3, T2-4)
\section{The bounded-learner gap}
\label{sec:bounded}

% Area role (charter §2): where format can help once information is held
% constant; an UPPER BOUND on achievable representation advantage via
% capacity/sample complexity; the regime that activates it.
%
% This file holds:
%   - T1-3 (WOV-264): statistical-learning-theory foundations (PAC / agnostic
%     setup, VC dimension, Rademacher complexity, uniform-convergence bounds),
%     how a representation induces a hypothesis class H∘φ and changes its
%     complexity at fixed information, a sample-complexity SEPARATION between
%     information-equivalent representations, and identification of the single
%     quantity that upper-bounds the achievable advantage.
%   - T2-4: the advantage upper bound itself (placeholder retained below).
%
% The positive content is delivered as falsifiable predictions in §8, never as
% established findings (guardrail G2).

\Cref{sec:parity} closes the information channel: under parity the Bayes risk
does not move with the representation (\cref{thm:irrelevance}, proved in T2-2).
If a representation effect exists at all, it must therefore live in the gap
between the Bayes-optimal predictor and a real, \emph{bounded} one --- finite
capacity, finite sample, bounded compute. This section builds the
learning-theoretic tools that measure that gap and identifies the single
quantity that will upper-bound the achievable advantage in T2-4. The framing
is deliberately one-directional: the theory bounds how large a parity-preserving
representation effect \emph{can} be, and names the regime in which to look for
it; it does not assert that any such effect exists (guardrail~G2). We follow
\citet{shalevshwartz2014} and \citet{mohri2018} for the learning theory,
\citet{vapnik1998} for the VC framework, and \citet{bartlettmendelson2002} for
the Rademacher bounds.

\subsection{The agnostic learning setup}
\label{subsec:pac}

Fix the data law $\jlaw$ of \cref{ass:sbs} and a representation $\rep$. A
learner sees $\nsample$ i.i.d.\ draws $S = \{(X_i,Y_i)\}_{i=1}^{\nsample}$
from $\jlaw$ and selects a predictor from a \emph{hypothesis class} of
$\rep$-measurable rules. Because a predictor on $\rep$ acts only through
$\rep(X)$ (\cref{subsec:pred}), it is natural to fix a base class
$\hypclass$ of maps $\Zspace_{\rep}\to\Aspace$ and let the representation
induce the working class by precomposition.

\begin{definition}[Induced hypothesis class]
\label{def:induced-class}
For a base class $\hypclass \subseteq \Aspace^{\Zspace_{\rep}}$ and a
representation $\rep$, the \emph{induced class} is
\[
  \hypclass \circ \rep \;\defeq\;
  \{\, h \circ \rep : h \in \hypclass \,\} \subseteq \Aspace^{\Xspace}.
\]
Its members are the $\repsig$-measurable rules the learner can express.
\end{definition}

\begin{definition}[Risk, empirical risk, excess risk]
\label{def:emp-risk}
With the loss $\loss$ of \cref{def:risk}, a hypothesis
$f = h\circ\rep$ has risk $\risk(f) = \Expect_{\jlaw}[\loss(Y,f(X))]$ and
empirical risk
$\emprisk(f) = \tfrac1{\nsample}\sum_{i=1}^{\nsample}\loss(Y_i,f(X_i))$. The
empirical risk minimizer (ERM) is
$\widehat{f}_{\rep} \in \arg\min_{f\in\hypclass\circ\rep}\emprisk(f)$. Its
\emph{excess risk} decomposes as
\begin{equation}
  \label{eq:excess}
  \risk(\widehat{f}_{\rep}) - \bayesrisk(\mathrm{id})
  \;=\;
  \underbrace{\Big(\inf_{f\in\hypclass\circ\rep}\risk(f) - \bayesrisk(\rep)\Big)}_{\text{approximation, w.r.t. }\rep}
  + \underbrace{\Big(\bayesrisk(\rep)-\bayesrisk(\mathrm{id})\Big)}_{\text{information gap (\cref{sec:parity})}}
  + \underbrace{\Big(\risk(\widehat{f}_{\rep}) - \inf_{f\in\hypclass\circ\rep}\risk(f)\Big)}_{\text{estimation}}.
\end{equation}
\end{definition}

The middle term in \eqref{eq:excess} is exactly the information channel of
\cref{sec:parity}: under parity it is zero and identical across $\rep,\rep'$
(\cref{thm:irrelevance}). The representation can therefore act on a bounded
learner only through the \emph{approximation} and \emph{estimation} terms ---
how well the base class composed with $\rep$ can approximate the optimum, and
how fast the ERM converges to the best in class. Both are controlled by the
\emph{capacity} of $\hypclass\circ\rep$, the object of the next two
subsections.

\subsection{Capacity: VC dimension and Rademacher complexity}
\label{subsec:capacity}

\begin{definition}[VC dimension]
\label{def:vc}
For a class $\mathcal{G}$ of $\{0,1\}$-valued functions on a domain
$\mathcal{D}$, a set $\{z_1,\dots,z_k\}\subseteq\mathcal{D}$ is
\emph{shattered} if $\mathcal{G}$ realizes all $2^k$ labelings of it. The
\emph{VC dimension} $\mathrm{VC}(\mathcal{G})$ is the size of the largest
shattered set (possibly $\infty$).
\end{definition}

\begin{definition}[Empirical Rademacher complexity]
\label{def:rademacher}
For a real-valued class $\mathcal{G}$ and a sample
$z_{1:\nsample}$, the \emph{empirical Rademacher complexity} is
\[
  \Rad{\mathcal{G}}
  \;\defeq\;
  \Expect_{\sigma}\Big[\;
    \sup_{g\in\mathcal{G}}\;
    \tfrac1{\nsample}\textstyle\sum_{i=1}^{\nsample}\sigma_i\,g(z_i)
  \;\Big],
\]
where $\sigma_1,\dots,\sigma_{\nsample}$ are i.i.d.\ Rademacher signs
($\Prob(\sigma_i=\pm1)=\tfrac12$). It measures how well $\mathcal{G}$
correlates with random noise --- its effective richness on the sample.
\end{definition}

We write $\capac{\mathcal{G}}$ for whichever capacity term is in force (the VC
dimension or the Rademacher complexity); the bounds below are stated for
$\Rad{\cdot}$, the sharper of the two, with the VC route available through the
standard $\Rad{\mathcal{G}} = O(\sqrt{\mathrm{VC}(\mathcal{G})/\nsample})$
chaining bound \citep[Ch.~6]{shalevshwartz2014}.

\subsection{Uniform convergence and the generalization bound}
\label{subsec:uniform}

\begin{theorem}[Rademacher uniform-convergence bound]
\label{thm:uniform}
Let $\loss$ take values in $[0,1]$ and let $\mathcal{F}=\hypclass\circ\rep$.
With probability at least $1-\delta$ over $S\sim\jlaw^{\otimes\nsample}$, every
$f\in\mathcal{F}$ satisfies
\begin{equation}
  \label{eq:uniform}
  \risk(f) \;\le\; \emprisk(f)
  \;+\; 2\,\Rad{\loss\circ\mathcal{F}}
  \;+\; \sqrt{\tfrac{\log(1/\delta)}{2\nsample}},
\end{equation}
where $\loss\circ\mathcal{F}$ is the induced loss class. Consequently the ERM
$\widehat{f}_{\rep}$ has estimation error
$\risk(\widehat{f}_{\rep}) - \inf_{f\in\mathcal{F}}\risk(f)
   \le 4\,\Rad{\loss\circ\mathcal{F}} + O(\sqrt{\log(1/\delta)/\nsample})$.
\end{theorem}

\begin{proof}[Proof sketch]
The two-sided uniform deviation $\sup_{f}|\risk(f)-\emprisk(f)|$ is bounded in
expectation by $2\,\Rad{\loss\circ\mathcal{F}}$ via symmetrization, and
concentrates around its mean by McDiarmid's inequality (the bounded-difference
constant is $1/\nsample$ since $\loss\in[0,1]$); the estimation-error
statement follows by applying \eqref{eq:uniform} to $\widehat{f}_{\rep}$ and to
the in-class optimum and adding. Full details:
\citet[Thm.~26.5]{shalevshwartz2014}, \citet[Ch.~3]{mohri2018}.
\end{proof}

\eqref{eq:uniform} is the formal statement that, at fixed sample size, a
\emph{lower-capacity} induced class generalizes from data with a smaller gap.
This is the lever a representation has on a bounded learner, and it is
entirely a statement about $\Rad{\loss\circ(\hypclass\circ\rep)}$ --- not about
information content.

\subsection{Representations change capacity at fixed information}
\label{subsec:rep-capacity}

The crux of the positive half of the thesis is that the two channels ---
information (\cref{sec:parity}) and capacity (this section) --- are
\emph{independent}. A representation can leave the information term
\eqref{eq:dpi-rep} untouched while moving the capacity term \eqref{eq:uniform}.

\begin{proposition}[Capacity is not an information invariant]
\label{prop:capacity-separation}
There exist a base class $\hypclass$ and two representations $\rep,\rep'$ in
information parity for $Y$ (\cref{def:parity}, so $\bayesrisk(\rep) =
\bayesrisk(\rep')$) with
\[
  \Rad{\loss\circ(\hypclass\circ\rep)}
  \;\ne\;
  \Rad{\loss\circ(\hypclass\circ\rep')}.
\]
\end{proposition}

\begin{proof}
Take $Y=\mathrm{sign}\langle w^\star, \psi(X)\rangle$ for a fixed feature map
$\psi$ and weight $w^\star$, so that the linear-in-$\psi$ class is Bayes for
$Y$. Let $\rep=\psi$ (features in which the target is linearly separable) and
let $\rep'=\Pi(\psi)$ for an information-preserving but nonlinear
re-coordinatization $\Pi$ (a bijection on the range of $\psi$, so
$\sigalg{\rep'}=\sigalg{\rep}$ and a fortiori $\rep\parityeq\rep'$). With
$\hypclass$ the class of linear separators, $\hypclass\circ\rep$ contains the
Bayes rule with bounded-norm weights, whereas $\hypclass\circ\rep'$ must
realize the same decision boundary as a curved set in the new coordinates,
inflating the weight norm and hence the Rademacher complexity of the induced
loss class \citep[Thm.~26.12]{shalevshwartz2014}. The information content is
identical by construction; the capacity required to express the optimum is
not.
\end{proof}

\cref{prop:capacity-separation} is the formal locus of ``format can help'':
two representations that are interchangeable for an infinite-capacity
predictor can present very different learning problems to a bounded one. It is
also the formal boundary of guardrail~G2 --- the proposition asserts only that
the capacity terms \emph{can} differ, the direction and magnitude of any
real-LLM effect being an empirical question routed to the predictions of
\cref{sec:predictions}.

\subsection{The advantage-bounding quantity}
\label{subsec:advantage-quantity}

Combining \eqref{eq:excess} under parity (information gap $=0$) with
\eqref{eq:uniform} isolates the quantity that controls everything.

\begin{definition}[Representation advantage]
\label{def:advantage}
For two parity representations $\rep\parityeq\rep'$ and a fixed bounded learner
(base class $\hypclass$, sample size $\nsample$), the \emph{representation
advantage} is the achieved excess-risk difference
\[
  \advantage(\rep,\rep')
  \;\defeq\;
  \big[\risk(\widehat{f}_{\rep'}) - \bayesrisk(\mathrm{id})\big]
  - \big[\risk(\widehat{f}_{\rep}) - \bayesrisk(\mathrm{id})\big].
\]
\end{definition}

Under parity the information-gap term in \eqref{eq:excess} cancels between the
two representations, so $\advantage(\rep,\rep')$ is built entirely from the
approximation and estimation terms, and \cref{thm:uniform} bounds the
estimation part by the induced-class Rademacher complexity. This identifies
the single quantity that will upper-bound the achievable advantage in T2-4:
\begin{equation}
  \label{eq:advantage-quantity}
  \boxed{\;\;
  \text{the induced-class capacity gap}\quad
  \Rad{\loss\circ(\hypclass\circ\rep)}
  \;\;\text{vs.}\;\;
  \Rad{\loss\circ(\hypclass\circ\rep')}
  \;\;}
\end{equation}
together with the (representation-dependent) approximation error. T2-4 turns
this into a one-sided bound $\advantage(\rep,\rep') \le \Phi(\capac{\cdot},
\nsample)$ controlled by capacity and sample size and \emph{not} by
information content --- the formal statement that any parity-preserving effect
is a bounded-computation phenomenon of bounded size (charter §1).

\begin{proposition}[Bounded-regime advantage is capacity-controlled --- placeholder]
\label{prop:advantage}
\emph{[Placeholder for T2-4.]} The achievable representation advantage
$\advantage(\rep,\rep')$ at finite capacity satisfies an upper bound in terms
of $\capac{\hypclass}$ (via the induced-class complexities
\eqref{eq:advantage-quantity}) and the sample size $\nsample$.
\idealization{Holds in the bounded regime (finite capacity / finite sample /
bounded compute); it does not assert any effect exists, only bounds its
maximum magnitude. The bound is stated under parity, so the information channel
of \cref{sec:parity} contributes nothing to it.}
\end{proposition}
