% Section 10 — Limitations and honesty
% Integrity gate: adds no new result; audits the ones already proved and states
% their price.
\section{Limitations and honesty}
\label{sec:limitations}

This section is the manuscript's integrity gate. A math-heavy account built on
top of a null is most at risk of reading
as obfuscation, and the antidote is to collect in one place three things that
are otherwise distributed through the paper: the assumptions every result rests
on and the way each fails for a real frozen LLM (\cref{subsec:lim-assumptions});
an explicit list of what the theory does \emph{not} establish
(\cref{subsec:lim-notestablish}); and the record of an internal red-team pass
that re-read the whole draft for two specific failure patterns --- converting
the null into a positive structure claim, and presenting an idealized theorem as
an empirical fact (\cref{subsec:lim-redteam}). Nothing here is a new theorem; the
section audits what the previous sections proved and names what it costs.

\subsection{Assumptions and their failure modes}
\label{subsec:lim-assumptions}

The results stand on six assumptions. Two are the idealizations disclosed at
every use through the \texttt{\textbackslash idealization} boxes
and named in \cref{rem:slogan}: \emph{(i)~Bayes-optimality} and \emph{(ii)~exact
parity}. The other four are modeling choices made once --- in \cref{sec:prelim}
and \cref{sec:bounded} --- and inherited everywhere downstream.
\cref{tab:idealization-ledger} is the ledger: each assumption, where it enters,
where (if anywhere) it is relaxed or bounded rather than merely assumed, and the
failure mode it carries for the real apparatus. Collecting them discharges the
obligation that no single result silently accumulates more idealization than
it states (\cref{rem:parity-ledger}).

\begin{table}[t]
\centering
\small
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.10\linewidth}>{\raggedright\arraybackslash}p{0.22\linewidth}>{\raggedright\arraybackslash}p{0.24\linewidth}>{\raggedright\arraybackslash}p{0.30\linewidth}@{}}
\toprule
Assumption & Enters at & Relaxed / bounded at & Failure mode for a frozen LLM \\
\midrule
\textbf{A1} Bayes-optimality \emph{(ideal.~(i))} &
  \cref{thm:irrelevance}, \cref{cor:ratedist-gap}, all of \cref{sec:anchoring} &
  \cref{sec:bounded} (the gap \emph{is} the relaxation); \cref{thm:defect-risk-gap} (a fixed bounded predictor) &
  The model is not the posterior $\post{\rep}$; the at-optimum equality is unobservable, and what is seen is the bounded residual $\le\Phi$, claimed small, never structurally zero. \\
\textbf{A2} Exact parity \emph{(ideal.~(ii))} &
  \cref{thm:irrelevance} via $\infogap{\cdot}=0$; every operational-parity row of \cref{tab:parity-ledger} &
  \cref{sec:parity-defect}: the residual is \emph{measured} as $\paritydefect$, not assumed away (\cref{thm:defect-risk-gap}) &
  $\ladder{4},\ladderplus$ are operationally but not absolutely parity (\cref{prop:abs-strict}); if the projection lost predictive information the gap would carry a rate--distortion term (\cref{cor:ratedist-gap}), so parity is re-verified per cell (\cref{sec:predictions}). \\
\textbf{A3} Kernel/risk model &
  \cref{sec:prelim}: representation $=$ measurable map, predictor $=$ Markov kernel, one decision loss with a Bayes act &
  Not relaxed; it is the frame. Loss taken bounded ($[0,1]$) for \cref{sec:bounded}, Lipschitz/bounded for \cref{sec:parity-defect} &
  An LLM is autoregressive and its ``loss'' is a scorer output, not a fixed decision loss; the constant $L$ and the metric $\Fisher$ of \cref{ex:r4-defect} are posited, not estimated. \\
\textbf{A4} i.i.d.\ sample / ERM analogy &
  \cref{sec:bounded}: $\nsample$ draws, base class $\hypclass$, ERM $\widehat f_{\rep}$ &
  Read onto the apparatus by analogy, flagged at each use (\cref{sec:anchoring} preamble) &
  A frozen model queried in-context is not an ERM over a tunable sample; the analogy may break, so \cref{sec:anchoring,sec:predictions} assert only bounds, regimes, and trends, never point values. \\
\textbf{A5} In-distribution scope &
  The frozen apparatus: one model, single-function non-recursive Python, the fixed C-ladder and scorers &
  Not relaxed; the theorems are domain-agnostic but the anchoring is not &
  Nothing licenses extrapolation to multi-file, recursive, cross-language, or agentic code; the predictions name knobs on \emph{this} apparatus (\cref{subsec:battery}). \\
\textbf{A6} Conditional section structure &
  \cref{sec:category}, \cref{sec:parity-defect}: each survives only by earning one theorem &
  Resolved (\cref{thm:retraction}; \cref{thm:defect-risk-gap}); demotion criteria still on record &
  The numeric defect of \cref{ex:r4-defect} is illustrative/computed-in-principle (\cref{rem:defect-status}); ``the control was near exact parity'' is a labeled conjecture. \\
\bottomrule
\end{tabular}
\caption{The assumption ledger. A1--A2 are the two idealizations of
\cref{rem:slogan}, disclosed at every use; A3--A5 are modeling choices inherited
from \cref{sec:prelim,sec:bounded}; A6 is the provisional status of the two
conditional sections. ``Relaxed / bounded'' distinguishes assumptions the paper
later \emph{discharges into a measurable quantity} (A1 into $\Phi$, A2 into
$\paritydefect$) from those it simply carries (A3--A5).}
\label{tab:idealization-ledger}
\end{table}

Three of these deserve a sentence beyond the table. \textbf{A1 and A2 are not
symmetric.} The paper does real work to retire each into something observable
rather than leaving it as an article of faith: Bayes-optimality (A1) is relaxed
by the entire bounded-learner apparatus, so the quantity that survives is the
capacity budget $\Phi$ (\cref{thm:bounded-gap}), not an appeal to an unreachable
optimum; exact parity (A2) is relaxed by the parity defect, so the residual is
the measurable $\paritydefect$ (\cref{thm:defect-risk-gap}), not an assumption.
What is \emph{not} retired is the joint claim that a single frozen model behaves
like the idealized objects at all --- that is A4, the analogy, and it is the
load-bearing soft spot of the empirical mapping, which is exactly why
\cref{sec:anchoring} and \cref{sec:predictions} restrict themselves to bounds,
prohibitions, and monotone trends. \textbf{A5 is a scope fence, not a defect of
the proofs.} The theorems of \cref{sec:parity,sec:bounded,sec:parity-defect} are
stated over an abstract standard Borel space and hold wherever their hypotheses
do; the limitation is that the only apparatus they have been \emph{anchored} to
is the narrow one of \texttt{preprint-v1}, and a reader should treat every
empirical sentence as scoped to it.

\subsection{What the theory does not establish}
\label{subsec:lim-notestablish}

The single most important honesty obligation is to be explicit about the gap
between what was proved and what a casual reading might infer. The following are
\emph{not} consequences of anything in this paper.

\begin{enumerate}
  \item[(N1)] \emph{It does not prove that representation never helps.}
    \cref{thm:irrelevance} equalizes the \emph{Bayes-optimal} risks under
    parity; it is silent on a bounded learner, for which \cref{thm:bounded-gap}
    permits a nonzero advantage up to $\Phi$. ``No information-channel effect is
    expected'' is the claim; ``the effect is zero'' is not, and conflating the
    two is the central overclaim this paper is at pains to avoid
    (\cref{rem:anchor-limits}, L2).
  \item[(N2)] \emph{It does not establish that any effect exists, or its sign.}
    Every bound is one-sided and sign-agnostic (\cref{thm:bounded-gap},
    \cref{cor:size-gap}): it limits how large a parity-preserving advantage
    \emph{can} be and is mute on whether one is realized and on which
    representation would win. The positive half is delivered only as falsifiable
    predictions, never as findings (\cref{rem:pred-honesty}).
  \item[(N3)] \emph{It does not confirm the theory from the frozen record.}
    \cref{sec:anchoring} claims consistency, not confirmation (L1): a null and a
    ceiling fail to refute, which is weaker than corroborating. The theory is put
    at risk only by the predictions of \cref{sec:predictions}, which are not yet
    collected.
  \item[(N4)] \emph{It does not compute the parity defect on the frozen
    predictor.} The numeric $\paritydefect\approx0.089$ of \cref{ex:r4-defect}
    is illustrative: it consumes the measured round-trip recovery rates but normalizes
    the action geometry, fixes $L=\lambda=1$, and posits a diagonal $\Fisher$
    rather than estimating one (\cref{rem:defect-status}). The statement ``the
    control was near exact parity'' is therefore a labeled conjecture until the
    recipe is run.
  \item[(N5)] \emph{It does not prove the T1/T3 regime distinction.} That T3
    sits below the capacity frontier --- hence realizable, hence saturated
    rather than information-limited --- is an empirical surmise about task
    difficulty, not a theorem (L4). Were T3 in fact frontier-hard,
    only the ceiling-censoring half of \cref{subsec:anchor-t3} would survive.
  \item[(N6)] \emph{It does not recommend building anything.} The decision-gate
    mapping of \cref{tab:pred-gates} is conditional on outcomes not yet observed;
    a confirmed bounded-regime effect would scope a product to that regime, a
    clean null would corroborate the negative half, and a non-decaying effect
    would refute the thesis. The paper supplies the test, not a verdict.
\end{enumerate}

\subsection{Internal red-team against overclaim}
\label{subsec:lim-redteam}

The integrity gate is not only a list of caveats; it is a discipline applied to
the text. We re-read the manuscript end to end --- abstract, \cref{sec:intro}
through \cref{sec:conclusion}, and the appendix --- at sentence granularity,
flagging any clause matching one of two patterns: \textbf{(P-a)} a sentence that
converts the recorded null into a positive structure claim or builds a
structure-advantage assertion on top of it, or
\textbf{(P-b)} a sentence that states a result holding only at an idealization
as though it held of the frozen LLM. \cref{tab:redteam}
records, for each integrity check, where it is enforced in-source and the verdict of
the pass.

\begin{table}[t]
\centering
\small
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.07\linewidth}>{\raggedright\arraybackslash}p{0.55\linewidth}>{\raggedright\arraybackslash}p{0.27\linewidth}@{}}
\toprule
Check & Where enforced in-source & Verdict \\
\midrule
Null not a shortfall & The null is presented as the predicted information-limit outcome
  (\cref{thm:irrelevance}, \cref{subsec:anchor-t1}); every ``disappointing/
  should-have-been-positive'' phrasing appears only inside its own negation
  (\cref{sec:intro}, \cref{sec:anchoring}). & Holds. No sentence frames the null
  as a shortfall. \\
Positive half one-sided & The positive half is a one-sided bound (\cref{thm:bounded-gap},
  \cref{cor:size-gap}); \cref{sec:predictions} is predictions/conjectures only
  (\cref{rem:pred-honesty}); sign of $\advantage$ left free throughout. &
  Holds, after RT-1 below. \\
Idealizations disclosed & \texttt{\textbackslash idealization} boxes on \cref{thm:irrelevance},
  \cref{cor:ratedist-gap}, \cref{thm:bounded-gap}, \cref{thm:defect-risk-gap},
  \cref{prop:defect-fisher-local}, \cref{subsec:anchor-t1}; collected in
  \cref{tab:idealization-ledger}. & Holds. Every invocation discloses (i)/(ii). \\
No decorative math & The two conditional sections carry explicit criteria
  (\cref{subsec:cat-gonogo}, \cref{subsec:defect-gonogo}) and survive only by a
  theorem. & Holds; see demotion record. \\
Citation hygiene & Citation hygiene restated in the \cref{sec:related} header and the bib
  header; each cite verified against a primary source at insertion. & Holds. \\
Conjectures labeled & Conjectural steps labeled: the T1/T3 difficulty surmise
  (\cref{subsec:anchor-t3}, L4), the computed-in-principle defect
  (\cref{rem:defect-status}, \cref{ex:r4-defect}). & Holds. \\
\bottomrule
\end{tabular}
\caption{Red-team checklist. The pass looked for two failure patterns (P-a:
null$\to$positive; P-b: idealization-as-fact) against six integrity checks. The
one substantive finding is RT-1.}
\label{tab:redteam}
\end{table}

\paragraph{Findings.}
The pass produced one substantive correction. \textbf{RT-1 (positive-half consistency).}
\cref{sec:intro} and \cref{sec:conclusion} originally described the
\cref{sec:predictions} battery as ``sign-aware'' predictions, whereas
\cref{sec:predictions} itself is emphatic that every prediction is
\emph{sign-agnostic} --- the only directional content is the trend of the
\emph{magnitude} $|\advantage|$ across a resource knob, never the sign of
$\advantage$ (\cref{subsec:pred-form}, \cref{rem:pred-honesty}). The word
``sign-aware'' could be read as the theory committing to a direction, which is
exactly the manufactured-positive failure the one-sided-bound discipline forbids; both occurrences were
changed to ``sign-agnostic,'' matching the usage already standard in
\cref{sec:intro} (the contributions paragraph), \cref{sec:bounded}, and
\cref{sec:predictions}. No instance of pattern P-a or P-b survived the pass
otherwise: the abstract's ``we prove a two-halved answer'' is accurate (a
theorem and an upper bound), the conditional sections' ``show/earn'' verbs attach to
proved propositions, and \cref{sec:anchoring} never upgrades consistency to
confirmation. The end-to-end read also surfaced one non-overclaim correctness
defect --- a grouped cross-reference in \cref{subsec:battery} that rendered as a
broken ``\texttt{??}'' marker --- which was fixed in the preamble; it is recorded
here only so the pass is reported in full.

\paragraph{Demotion record.}
Both provisional sections were resolved in favour of keeping them, and we state
the bar they cleared so the reader can audit that it was real rather than
aesthetic. \cref{sec:category} (category theory) survived because
\cref{thm:retraction} proves the round-trip defect is an idempotent lying in the
common kernel of the mutual information (\cref{prop:suff-iff-mi}) and the Bayes
risk (\cref{thm:irrelevance}) --- a statement neither of those functionals can
make, hence not a re-encoding of either (criterion \cref{subsec:cat-gonogo}).
\cref{sec:parity-defect} (OT/information geometry) survived because
\cref{thm:defect-risk-gap} shows $\paritydefect$ upper-bounds, and
\cref{cor:defect-finer-order} shows it strictly refines the information-gap
ordering on the parity class --- both prongs of criterion
\cref{subsec:defect-gonogo}, witnessed by \cref{ex:r4-defect}.
Had either produced only a restatement of \cref{sec:parity}, the criteria
required demotion to a single remark; that neither did is the reason both
sections stand.

\begin{remark}[The honest residue, restated]
\label{rem:lim-residue}
What survives every caveat above is a single falsifiable claim, and it is worth
isolating it from the residue it is not. \emph{The residue:} the two
idealizations A1--A2 (relaxed into $\Phi$ and $\paritydefect$ but never met
exactly by a frozen model), the ERM analogy A4 (carried, not discharged), the
in-distribution scope A5, and the conjectural T1/T3 split (N5). None is hidden;
each is disclosed at use and collected in
\cref{tab:idealization-ledger,tab:redteam}. \emph{The claim that survives:}
format cannot supply predictive information that the input does not contain
(\cref{thm:irrelevance}), so any genuine code-representation effect on a bounded
model is a capacity phenomenon of bounded size (\cref{thm:bounded-gap}),
appearing only in the finite-resource, capacity-separated regime the theory
names (\cref{prop:regime}) --- and that claim is one the apparatus can be pointed
at and made to refute (\cref{sec:predictions}). The limitations bound where the
claim applies; they do not soften it.
\idealization{This section introduces no new reliance on (i) or (ii); it is the
ledger of the reliances incurred elsewhere. The one place it makes a positive
assertion of its own --- that the residue is fully disclosed --- is verifiable
by cross-checking \cref{tab:idealization-ledger,tab:redteam} against the
\texttt{\textbackslash idealization} boxes of \cref{sec:parity,sec:bounded,%
sec:parity-defect,sec:anchoring}, which is the audit the disclosure discipline asks the reader
to be able to perform.}
\end{remark}
